The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2014-09-09 Papers/Universalis/ModalCollapse.tex
contributions to the paper

A crucial step of most ontological arguments is the claim that if
God's existence is possible, then it is necessary.  This is Lemma
\textbf{L2} in G\"odel's proof.  In the modal square of opposition
(Fig. \ref{fig:square}), this is an unusual situation in which the
\textbf{I} corner must imply and entail the \textbf{A} corner, in the
particular case when $\phi$ is $\exq x G(x)$.  G\"odel's proof shows
that his axioms are indeed strong enough to invert the direction of
entailment this choice of $\phi$.
% for the sentence at issue.  
This observation, however, immediately brings up the question whether
the axioms are even strong enough to enable the inverted entailment
for arbitrary sentences $\phi$.  That is essentially the question
asked by Sobel \cite{Sobel1987}, and his proof of the modal collapse
(\textbf{MC}, cf. Fig.~\ref{fig:collapse}) provides an affirmative answer. It is possible to show
that this form of the modal collapse entails (in modal logic
\textbf{\emph{K}}) a collapse of the modal square (\textbf{MC'}),
causing the subcontraries to entail (and even imply) their respective
contraries. Normally, as shown in Fig. \ref{fig:square}, in the modal
square of opposition only the other direction of entailment holds: the
contraries entail their subcontraries, assuming the \emph{modal
  existential import} \textbf{ExImp}
\cite{WhatToCiteHere}.\footnote{C.: We need a reference here; I am still
  searching for good papers; this here is also interesting:
  \url{http://www.tandfonline.com/doi/pdf/10.1080/01445340.2013.764962}}

@@

All results announced in this note have been obtained experimentally
using interactive and automated theorem provers and model finders
\cite{LEO,Satallax,Isabelle,Coq,Nitpick}.  The source codes of the
experiments, as well as the resulting proofs and counter-models, are
available in
\url{https://github.com/FormalTheology/GoedelGod/tree/master/Formalizations/Isabelle}.\footnote{Here
  we should point to the exact files (and put them in a relatively
  stable subdirectory?). We also should also use corresponding names in
  these files, so that the reader easily finds the related 
  parst. Right now this is still quite difficult.}

>>>>

A crucial step of most ontological arguments is the claim that if
God's existence is possible, then it is necessary.  This is Lemma
\textbf{L2} in G\"odel's proof.  In the modal square of opposition
(Fig. \ref{fig:square}), this is an unusual situation in which the
\textbf{I} corner must imply and entail the \textbf{A} corner, in the
particular case when $\phi$ is $\exq x G(x)$.  G\"odel's proof shows
that his axioms are indeed strong enough to invert the direction of
entailment this choice of $\phi$.
% for the sentence at issue.  
This observation, however, immediately leads to the question whether
the axioms are eventually even strong enough to enable the inverted entailment
for arbitrary sentences $\phi$.  That is essentially the question
asked by Sobel \cite{Sobel1987}, and his proof of the modal collapse
(\textbf{MC}, cf. Fig.~\ref{fig:collapse}) provides an affirmative answer. It is possible to show
that this form of the modal collapse entails (in modal logic
\textbf{\emph{K}}) a collapse of the modal square (\textbf{MC'}),
causing the subcontraries to entail (and even imply) their respective
contraries. Normally, as shown in Fig. \ref{fig:square}, in the modal
square of opposition only the other direction of entailment holds: the
contraries entail their subcontraries, assuming the \emph{modal
  existential import} \textbf{ExImp}
\cite{WhatToCiteHere}.\footnote{C.: We need a reference here; I am still
  searching for good papers; this here is also interesting:
  \url{http://www.tandfonline.com/doi/pdf/10.1080/01445340.2013.764962}}

@@

All results announced in this note have been obtained experimentally
using interactive and automated theorem provers and model finders
\cite{LEO,Satallax,Isabelle,Coq,Nitpick}.  The source codes of the
experiments, as well as the resulting proofs and counter-models, are
available in
\url{https://github.com/FormalTheology/GoedelGod/tree/master/Formalizations/Isabelle}.\footnote{C.: Here
  we should point to the exact files (and put them in a relatively
  stable subdirectory?). We also should also use corresponding names in
  these files, so that the reader easily finds the related 
  parst. Right now this is still quite difficult.}
